---
title: "Statewide Groundwater Resource Deletion Summary"
author: "Sage Davis"
date: "11/9/2021"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, comment = FALSE)
library(stats)
library(tidyverse)
library(latex2exp)
```

## Analysis 

### Approach Summary:
I performed a detailed analysis of the publication, "Visualizing A Stochastic Model of California Water Resources Using Sankey Diagrams" (Curmi et al. 2013), where I gathered the groundwater inputs and changes for 2000 and the projections for 2050.  Using these values I created a simple net mass balance for the net change in groundwater at year 2000 and 2050. which equals the inputs minus the outputs.  With this model I was able to calculate the inputs for 2000 and 2050.  I used the data analysis software called "R" to perform my analysis.  Individual linear models (regressions) for the groundwater inputs and outputs between years 2000 to 2050. The results of the regression was used to calculate the change in the groundwater storage each year. 
  Using the values on Figure 7 (Curmi et al. 2013), of an approximate average groundwater resource in 2000 of *350x10^9 m3*. I made an educated assumption that the uncertainty exhibits a normal distribution around that mean value with a *115x10^9 m3* standard distribution.  Using the mean and standard deviation on a normal distribution, there is a 90% chance that the that the actual amount of California groundwater lies between the approximate lower limit of *190x10^9 m3* and approximate upper limit of *550x10^9 m3*
  






```{r echo = FALSE, warning = FALSE, comment = FALSE}
# 1. Set up a simple mass balance model of groundwater driven by inputs and outputs.


# 2. Specify the initial inputs and outputs for 2000. This requires determining the inputs and losses of groundwater in Figure 2.


# In the year 2000, according to Figure 2.

out00 = 18.2             # Total groundwater use (output)  
chng00 = -5.4             # Change in groundwater storage (input)
in00 =  out00 + chng00     # Net change in groundwater
#in00

```



```{r echo = FALSE, warning = FALSE, comment = FALSE}
#3. Determine the conditions at the end of the simulation (2050). This requires determining the inputs and outputs of groundwater in Figure 4.

# In the year 2050, according to Figure 4.

out50 = 27               # Total groundwater use (output)  
chng50 = -16.7            # Change in groundwater storage (input)
in50 = out50 + chng50      # Net change in groundwater 
#in50

```



```{r echo = FALSE, warning = FALSE, comment = FALSE}

# 4. Fit a linear trend in the input and output terms between the year 2000 and 2050 in order to determine the transient changes in recharge and withdrawals over this 50-year period.

# fit linear trend lines between the years 2000-2050
year <- c(2000, 2050)

# trend line for inputs
in_trend <- c(in00, in50)
in_lm <- lm(in_trend ~ year)
#in_lm

# trend line for outputs
out_trend <- c(out00, out50)
out_lm <- lm(out_trend ~ year)
#out_lm
```

```{r echo = FALSE, warning = FALSE, comment = FALSE}
# create a dataframe with all the years from 2000-2050
years <- seq(from=2000, to=2050, by=1)
massbal <- as.data.frame(years)

# use the equation from lm to fill in columns for ins and outs
massbal$ins = (-0.05)*massbal$years + 112.8
massbal$outs = (0.176)*massbal$years - 333.8

# head(massbal)
```


```{r echo = FALSE, warning = FALSE, comment = FALSE, fig.keep = "all", fig.align = "center"}
# plot the ins and outs over the years - what will happen to storage based on these trends?
ggplot(massbal) + 
  geom_line(aes(x=years, y=ins, col='Input')) +
  geom_line(aes(x=years, y=outs, col='Output')) +
  xlab("Year") + 
  ylab(TeX("Groundwater Flux ($10^9$ $m^3$)")) + 
  ggtitle(TeX("CA Groundwater Flux Input/Output per Year, 2000 - 2025")) +
  theme(plot.title = element_text(hjust = 0.5))



```

**Figure 1.**  Groundwater Flux vs Year for 2000 - 2050 for groundwater input and outputs.  These are derived from two independent linear (regression) models assuming the relationship between 2000 and 2050 is linear.  The inputs are shown to decrease as the outputs increase over the time period.  
.  
```{r echo = FALSE, warning = FALSE, comment = FALSE, fig.keep = "all", fig.align = "center"}
# can do mass balance for each year to see the changes 
massbal$chnge = massbal$ins - massbal$outs
ggplot(massbal) + 
  geom_line(aes(x=years, y=chnge)) +
  xlab("Year") + 
  ylab(TeX("Change in Groundwater Storage ($10^9$ $m^3$)")) +
  ggtitle(TeX("CA Groundwater Storage per Year, 2000 - 2025")) +
  theme(plot.title = element_text(hjust = 0.5))

# linear model of the change each year
 massbal_chnge_lm <- lm(massbal$chnge ~ years)
```

**Figure 2.**  The change in California Groundwater Storage vs Year for 2000 - 2050.  This is the result of groundwater inputs minus outputs as shown in Figure 1.  This is derived from a linear (regression) model assuming the relationship between 2000 and 2050 is linear.  The groundwater storage change is becoming more negative, or trending downward, if California water use continues like business as usual.

```{r echo = FALSE, warning = FALSE, comment = FALSE}

# 5. Calculate the changes in groundwater resources from 2000-2050 for at least three scenarios based on varying initial conditions of groundwater availability: The expected amount of groundwater (from section 2, above) as well as for upper and lower estimates provided in section 2.

# - From Curmi et al., Figure 7, assume that the expected value of groundwater resources in 2000 was *350x10^9 m3*. Let's further assume that uncertainty regarding this value follows a normal distribution around that mean value, with a standard deviation of 115x109 m3. 

# - Based on a normal distribution with this mean and standard deviation, there is a 90% chance that the actual amount of groundwater in CA is somewhere between the approximate lower limit of *190x10^9 m3* and an approximate upper limit of *550x10^9 m3*.

# all of the flows were also 10^9, so there is no need for unit conversions

# integrate over the water lost starting in 2000
water_lost <- function(year){
  int <- integrate(function(x){-0.226*x + 446.6}, lower = 2000, upper = year) # this function comes from the linear model of change over year
  return(int$value)
}
massbal$w_lost <- sapply(massbal$years, water_lost)

# create columns for the average, lower bounds, and upper bouds
massbal$gw_storage = massbal$w_lost + 350
massbal$gw_low = massbal$w_lost + 190
massbal$gw_upp = massbal$w_lost + 550

# head(massbal)

```



```{r echo = FALSE, warning = FALSE, comment = FALSE, fig.keep = "last", fig.align = "center"}
# use ggplot here to create a beautiful plot to include

ggplot(data = massbal) + 
  geom_rect(aes(ymin=-Inf, ymax=0, xmin=2000, xmax=2050), alpha = 0.01, fill='orange') +
  geom_line(aes(x = years, y = gw_storage, col = 'Average'), size = 1) +
  geom_line(aes(x = years, y = gw_low, col = 'Lower Bounds'), linetype='dashed', size = 1) + 
  geom_line(aes(x = years, y = gw_upp, col = 'Upper Bounds'), linetype='dashed', size = 1) +
  xlab("Year") + 
  ylab(TeX("Groundwater Stoarge ($10^9$ $m^3$)")) + 
  theme(legend.title = element_blank()) + 
  theme(panel.background = element_rect(fill = "light blue")) +
  ggtitle("Projected CA Groundwater Storage with 90% Confidence Interval")
  

```

**Figure 3.** The plot shows that if we use CA groundwater water like business as usual, with a 90% confidence interval.  California will run out water in 2024, 2037 or 2050 depending on 90% confidence level. The orange rectangle represents a negative water resource, no groundwater.  It depends on how much water storage exists in California.  This plots is derived by taking the CA groundwater change variable in figure 2, and performing a linear model with groundwater change as the dependent variable of years.  The change for each year is subtracted from the low, mean, and high approximations of the 2000 groundwater resource. I assumed the relationship is linear for this model to hold true. Using the year 2000 groundwater resource mean of *350x10^9 m3* (Curmi et al. 2013) with an assumed normal distribution, yields a standard distribution of *115x10^9 m3*.  Using the mean and standard deviation on a normal distribution, there is a 90% chance that the that the actual amount of California groundwater lies between the approximate lower limit of *190x10^9 m3* and approximate upper limit of *550x10^9 m3*.

### Discussion
California is running out of ground water.  We need to be strategist ways to decrease our water use so that the rate of change can at least not continue increase over time.

More recent data would be helpful in fining the mass balance model.  Ths would allow us to verify or modify the model depending on the more current data.  I would suggest attempting to collect data on groundwater out and change every 5 years at a minimum.

### Citation
Curmi, E., Fenner, R., Richards, K. et al. Visualising a Stochastic Model of Californian Water Resources Using Sankey Diagrams. Water Resources Management 27, 3035–3050 (2013). https://doi.org/10.1007/s11269-013-0331-2


